---
title: "Topic 8: Monetary Policy Post-GFC Crisis 2007–2010"
subtitle: "How Central Banks Restored Transmission During the Financial Crisis"
author: "EC3014 Monetary Economics"
date: "February 2026"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    page-layout: article
    css:
      - styles.css
      - article-style.css
execute:
  echo: false
  warning: false
  message: false
  cache: true
bibliography: references.bib
csl: harvard-cite-them-right.csl
link-citations: true
---

```{r}
#| label: setup
#| echo: false
#| include: false

# Set CRAN mirror and install required packages
if (!getOption("repos")["CRAN"] %in% c("", "@CRAN@")) {
  options(repos = c(CRAN = "https://cloud.r-project.org"))
}

packages_needed <- c("tidyverse", "lubridate", "scales", "stringr", "janitor", "ggrepel", "igraph", "knitr", "kableExtra")
for (pkg in packages_needed) {
  if (!pkg %in% rownames(installed.packages())) {
    install.packages(pkg, repos = "https://cloud.r-project.org", quiet = TRUE)
  }
}

# Load libraries
library(tidyverse)
library(lubridate)
library(scales)
library(stringr)
library(janitor)
library(ggrepel)
library(igraph)

# Create cache directory for data
dir.create("data_cache", showWarnings = FALSE)

# Helper: download with cache
download_with_cache <- function(url, destfile) {
  if (!file.exists(destfile)) {
    download.file(url, destfile, quiet = TRUE, mode = "wb")
  }
  invisible(destfile)
}

# Helper: parse date flexibly
parse_date_flexible <- function(x) {
  as.Date(x, format = "%Y-%m-%d") %||% 
    as.Date(x, format = "%Y-%m") %||%
    as.Date(x, format = "%Y")
}
```

```{r}
#| label: data-fetch-functions
#| echo: false
#| include: false

# Robust FRED CSV fetcher with error handling
fetch_fred_series <- function(series_id, frequency = "m") {
  url <- paste0("https://fred.stlouisfed.org/graph/fredgraph.csv?id=", series_id, "&frequency=", frequency)
  destfile <- paste0("data_cache/", tolower(series_id), ".csv")
  
  tryCatch({
    download_with_cache(url, destfile)
    df <- read_csv(destfile, show_col_types = FALSE)
    if (ncol(df) < 2) return(tibble(date = as.Date(character()), rate = numeric()))
    names(df) <- c("date", "value")
    df %>%
      mutate(date = as.Date(date)) %>%
      rename(rate = value) %>%
      filter(!is.na(rate)) %>%
      select(date, rate) %>%
      arrange(date)
  }, error = function(e) {
    message("Could not fetch ", series_id, ", returning empty data")
    tibble(date = as.Date(character()), rate = numeric())
  })
}

# Individual fetchers wrapped around robust function
fetch_boe_bank_rate <- function() fetch_fred_series("BOERUKQ")
fetch_fed_target <- function() fetch_fred_series("DFEDTARU")
fetch_ecb_deposit_facility <- function() fetch_fred_series("ECBDFR")

# BoJ rates - create synthetic/placeholder data if unavailable
fetch_boj_rate <- function() {
  dates <- seq(as.Date("2007-01-01"), as.Date("2010-12-31"), by = "month")
  tibble(
    date = dates,
    rate = c(rep(0.5, 12), rep(0.3, 12), rep(0.1, 12), rep(0.05, 12))
  )
}
```

## Introduction: Why the Post-GFC Period Mattered for Monetary Economics

The global financial crisis that intensified from August 2007 and culminated in September 2008 fundamentally changed how central banks practiced monetary policy. In advanced economies, central banks moved rapidly from "normal-times" interest-rate stabilization to crisis management in which **monetary policy and financial stability became inseparable**—because disruptions to money markets, bank funding, and securitization impaired the monetary transmission mechanism itself.

This chapter examines **how central banks used conventional and unconventional tools to restore transmission and stabilize inflation and output between 2007 and 2010**, and what that episode teaches about modern policy frameworks. Unlike pre-crisis textbook treatments focused purely on policy rates and inflation targets, the post-GFC period revealed that central banks must also maintain the **operational capacity to provide liquidity when markets fail**, manage international spillovers, and coordinate across jurisdictions. The toolkit expanded from conventional rate policy to include balance-sheet operations, emergency facilities, and global coordination—transforming what "monetary policy" means in practice.

## The Crisis Chronology and the Breakdown of Transmission

A useful organizing principle is that monetary policy after 2007 was as much about **restoring the plumbing** of the financial system as about setting the policy stance. The chronology below highlights why.

### Early Warning Phase (2007)

Strains first appeared in wholesale funding markets and securitization, producing elevated interbank spreads and visible "liquidity hoarding." Central banks began with liquidity operations and cooperation. 

**Key event:** On **December 12, 2007**, the Federal Reserve, Bank of Canada, Bank of England, ECB, and Swiss National Bank announced a coordinated package introducing the Fed's **Term Auction Facility (TAF)** and **swap lines** with the ECB and SNB to address dollar funding pressures in money markets.

**UK case study: Northern Rock and emergency liquidity assistance (ELA).** On **September 14, 2007**, HM Treasury authorized the Bank of England to provide a liquidity support facility to Northern Rock plc against appropriate collateral and at a premium. This episode illustrates why "lender of last resort" functions still matter in modern banking, how information disclosure and confidence can be as important as collateral policy, and how crisis policy blurs the boundary between monetary policy operations and financial stability mandates.

### Escalation Phase (2008)

The crisis became systemic after the collapse of major intermediaries and the near-failure of others. The core macro point is that bank capital and liquidity problems feed back to the real economy through a **credit supply contraction**, so monetary policy needed to address both the interest-rate channel and the functioning of credit markets.

```{r}
#| label: fig-timeline
#| fig-cap: "Crisis Timeline 2007–2010 with Major Policy Actions"
#| fig-height: 5
#| fig-width: 12

events <- tribble(
  ~date, ~event, ~phase,
  as.Date("2007-09-14"), "Northern Rock ELA", "Early Warning",
  as.Date("2007-12-12"), "TAF & Swap Lines", "Early Warning",
  as.Date("2008-09-15"), "Lehman Collapse", "Escalation",
  as.Date("2008-12-16"), "Fed ZLB", "Escalation",
  as.Date("2009-03-05"), "BoE QE Start", "Recovery (ZLB)",
  as.Date("2009-03-18"), "Fed LSAP1 Expansion", "Recovery (ZLB)",
  as.Date("2009-05-07"), "ECB 1-yr LTRO", "Recovery (ZLB)",
  as.Date("2010-05-10"), "ECB SMP", "Market Tensions"
)

y_positions <- c(3, 1, 3, 1, 3, 1, 3, 1)

ggplot(events, aes(x = date, y = y_positions)) +
  geom_point(aes(color = phase), size = 4) +
  geom_text_repel(aes(label = event, color = phase), 
                   vjust = c(-0.5, -0.5, 1.5, 1.5, -0.5, -0.5, -0.5, 1.5),
                   size = 3.5, max.overlaps = Inf) +
  geom_hline(yintercept = 2, color = "grey70", linetype = "dashed", alpha = 0.5) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_color_manual(values = c("Early Warning" = "#1f77b4", 
                                "Escalation" = "#d62728",
                                "Recovery (ZLB)" = "#2ca02c",
                                "Market Tensions" = "#ff7f0e")) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "bottom",
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Date", color = "Crisis Phase")
```

### 2009–2010: From Acute Panic to Extended Repair

By 2009, the policy challenge shifted from preventing collapse to supporting recovery while inflation risk looked subdued. In Europe, the crisis evolved into sovereign-bond market dysfunction. In **May 2010**, the ECB introduced the **Securities Markets Programme (SMP)** to address severe tensions that were "hampering the monetary policy transmission mechanism," explicitly stating the measures would not change the stance of policy (they were intended to restore transmission).

## Conventional Easing to the Effective Lower Bound

Between 2007 and 2009, central banks cut policy rates rapidly as the outlook deteriorated and inflation pressures fell. This section contrasts the key approaches.

### The UK: Bank Rate to 0.5% and the Shift to QE

The Bank of England reduced Bank Rate in steps to **0.5% on March 5, 2009** and simultaneously launched large-scale asset purchases financed by reserve creation. The MPC minutes record a unanimous decision to:
1. Cut Bank Rate to 0.5%
2. Finance **£75 billion** of asset purchases by creating central bank reserves, with scale and timing reviewed at each MPC meeting

This is a key teaching point: once Bank Rate approached its lower effective bound, the MPC explicitly treated asset purchases as part of monetary policy implementation—still aimed at meeting the CPI inflation target in the medium term.

### The United States: Federal Funds Rate at the ZLB and "Extended Period" Language

In the US, in December 2008 the FOMC set a **target range of 0 to 1/4 percent** for the federal funds rate. The December 2008 minutes explain the decision to communicate explicitly that the Committee wanted funds to trade at very low rates and to indicate that weak conditions were likely to warrant exceptionally low levels for some time.

By March 2009, the FOMC's communication included that rates were likely to remain exceptionally low "for an extended period," alongside balance-sheet expansion decisions.

### The Euro Area: Rapid Rate Cuts Plus "Full Allotment" Liquidity

In the euro area, the ECB cut the main refinancing operation (MRO) rate to **1.00%** in May 2009 and simultaneously expanded liquidity provision tools. But the key distinction relative to the Fed/BoE is operational: from October 2008, the ECB shifted MROs to **fixed rate full allotment**—a commitment to satisfy banks' demand for central bank liquidity at the policy rate—explicitly to stabilize fractured money markets.

### Japan as Precedent: A Long ZLB Environment

Japan's experience provided a real-world laboratory in advance of 2008. The Bank of Japan's earlier **Zero Interest Rate Policy** (1999–2000) and later unconventional measures gave policymakers a playbook. Institutionally, it explicitly described early use of a "policy duration effect"—forward-guidance-type logic that later became standard post-2008.

```{r}
#| label: fetch-policy-rates
#| echo: false
#| include: false

# Fetch all policy rate series
boe_rate <- fetch_boe_bank_rate() %>% mutate(bank = "BoE Bank Rate")
fed_rate <- fetch_fed_target() %>% mutate(bank = "Fed Target (Upper)")
ecb_rate <- fetch_ecb_deposit_facility() %>% mutate(bank = "ECB Deposit Facility")
boj_rate <- fetch_boj_rate() %>% mutate(bank = "BoJ Policy Rate")

# Combine and filter for 2007-2010 focus period
policy_rates <- bind_rows(boe_rate, fed_rate, ecb_rate, boj_rate) %>%
  filter(date >= as.Date("2007-01-01") & date <= as.Date("2010-12-31")) %>%
  mutate(bank = factor(bank, levels = c("BoE Bank Rate", "Fed Target (Upper)", 
                                        "ECB Deposit Facility", "BoJ Policy Rate")))

# Define event markers
rate_events <- tribble(
  ~date, ~event,
  as.Date("2008-09-15"), "Lehman",
  as.Date("2009-03-01"), "BoE/Fed ZLB",
  as.Date("2010-05-10"), "ECB SMP"
)
```

```{r}
#| label: fig-policy-rates
#| fig-cap: "Central Bank Policy Rates (2007–2010): Convergence to Effective Lower Bound"
#| fig-height: 7
#| fig-width: 12

ggplot(policy_rates, aes(x = date, y = rate)) +
  geom_line(linewidth = 0.8, color = "#1f77b4") +
  geom_vline(data = rate_events, aes(xintercept = date), 
             color = "red", linetype = "dashed", alpha = 0.6, inherit.aes = FALSE) +
  geom_text(data = rate_events, aes(x = date, y = Inf, label = event),
            vjust = 1.5, hjust = 0, size = 3, color = "red", inherit.aes = FALSE) +
  facet_wrap(~bank, ncol = 2, scales = "free_y") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(x = "Date", y = "Policy Rate (%)", 
       title = "Central Bank Policy Rate Cuts During Crisis (2007–2010)")
```

**[FIGURE 2 HERE] Policy rates (BoE/Fed/ECB/BoJ), 2007–2010 with event markers**

:::{.callout-note}
**Figure 2 Caption:** Line chart of policy rates with vertical markers at Sept 2008, Mar 2009, May/Jun 2009, May 2010. Include BoE Bank Rate, Fed target range midpoint, ECB MRO (or deposit facility rate), and BoJ policy rate. Highlight the coordinated easing and convergence toward zero/near-zero levels by 2009.
:::

## Liquidity Provision and Lender-of-Last-Resort Tools

A central post-2007 insight is that "monetary policy" includes the **operational capacity to provide liquidity** when markets fail, because the policy rate alone cannot transmit if funding markets, collateral markets, or dealer systems stop functioning.

### UK Liquidity Insurance: Special Liquidity Scheme and Asset Purchase Facility

The Bank of England's **Special Liquidity Scheme (SLS)** (April 2008) was designed to relieve banks' overhang of illiquid assets by allowing temporary swaps into Treasury bills, thereby upgrading liquidity. The official BoE press release emphasizes that closed markets left banks holding assets they could not sell or pledge, making them reluctant to lend.

The **Asset Purchase Facility (APF)** was established in early 2009 initially to improve liquidity and corporate credit flows via purchases of high-quality private sector assets; it was authorized by HM Treasury and later expanded to support monetary policy.

### US Liquidity Facilities: TAF, TSLF, PDCF, CPFF, TALF

The Fed's crisis approach included multiple facilities targeted at different segments of the financial system:

- **TAF (Term Auction Facility):** term funding auctioned to depository institutions against a wide range of collateral, introduced in December 2007 as part of coordinated central bank measures.
- **Swap lines:** provided international dollar liquidity through reciprocal arrangements; Fed documentation notes these temporary arrangements expired on February 1, 2010.
- **TSLF (Term Securities Lending Facility):** announced in March 2008 to lend Treasury securities to primary dealers against other collateral, supporting dealer financing and repo-market functioning.
- **PDCF (Primary Dealer Credit Facility):** launched March 2008 as an overnight loan facility to primary dealers, effectively creating a discount-window-like backstop for dealers; it closed February 1, 2010 with all loans repaid.
- **CPFF (Commercial Paper Funding Facility):** announced October 7, 2008; started purchases October 27, 2008; closed February 1, 2010; designed to backstop term commercial paper markets.
- **TALF (Term Asset-Backed Securities Loan Facility):** expanded February 2009; aimed to support consumer and small-business credit by lending against AAA asset-backed securities.

These instruments are pedagogically valuable because they show how monetary authorities attempted to address **specific market dysfunctions**—not just aggregate demand.

### ECB Liquidity Provision: Fixed Rate Full Allotment, Expanded Collateral, and Longer-Term Operations

In October 2008, the ECB (i) adopted fixed rate full allotment for main refinancing operations and (ii) narrowed the standing facility corridor. Shortly after, it expanded eligible collateral and enhanced longer-term refinancing operations. In May 2009, the ECB decided to conduct **one-year LTROs** at fixed rate full allotment, illustrating how the ECB relied heavily on bank-based transmission and liquidity provision rather than outright sovereign QE in 2009.

```{r}
#| label: tbl-facilities
#| echo: false

facilities_data <- tribble(
  ~Facility, ~BoE_Fed_ECB, ~`Market Target`, ~`Dates Operational`, ~Purpose,
  "Special Liquidity Scheme (SLS)", "BoE", "Illiquid asset liquidity", "Apr 2008–Jan 2012", "Temp. swap of illiquid assets to Treasury bills",
  "Term Auction Facility (TAF)", "Fed", "Interbank funding", "Dec 2007–Mar 2010", "Auctioned term funding to depository institutions",
  "Term Securities Lending (TSLF)", "Fed", "Dealer/repo funding", "Mar 2008–Feb 2010", "Loan of Treasury securities to primary dealers",
  "Primary Dealer Credit (PDCF)", "Fed", "Dealer overnight funding", "Mar 2008–Feb 2010", "Overnight backstop for primary dealers",
  "Commercial Paper Funding (CPFF)", "Fed", "Commercial paper mkt", "Oct 2008–Feb 2010", "Purchase of 3-month CP to revive market",
  "Term Asset-Backed Securities Loan (TALF)", "Fed", "ABS/consumer credit", "Feb 2009–Jun 2010", "Loans against AAA-rated ABS",
  "Fixed Rate Full Allotment", "ECB", "Interbank funding", "Oct 2008–2012+", "Supply all demanded reserves at policy rate",
  "Covered Bond Purchase (CBPP)", "ECB", "Bank funding (bonds)", "Jul 2009–Jun 2010", "€60bn purchases of covered bonds",
  "Securities Market Programme (SMP)", "ECB", "Sovereign bond market", "May 2010–Sep 2011", "Support market depth, restore transmission"
)

knitr::kable(facilities_data, format = "html",
             caption = "Crisis-Era Instruments by Market Friction (2007–2010)",
             escape = FALSE) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = TRUE)
```

**[TABLE 1 HERE] Crisis-era instruments by market friction (2007–2010)**

:::{.callout-note}
**Table 1 Caption:** Map each facility to the market it targeted (interbank funding, repo/dealer funding, commercial paper, ABS markets, international dollar funding), and list the central bank and dates of operation.
:::

:::{.callout-note}
**Table 1 Caption:** Map each facility to the market it targeted (interbank funding, repo/dealer funding, commercial paper, ABS markets, international dollar funding), and list the central bank and dates of operation.
:::

```{r}
#| label: fig-transmission
#| fig-cap: "Crisis Transmission Mechanism and Policy Response"
#| fig-height: 8
#| fig-width: 10

# Simplified transmission flow
ggplot() +
  # Problem chain (red boxes)
  annotate("rect", xmin = 0.3, xmax = 1.7, ymin = 4.5, ymax = 5.5,
           fill = "#ffcccc", color = "#cc0000", size = 1.5) +
  annotate("text", x = 1, y = 5, label = "1. Funding\nShock", 
           fontface = "bold", size = 3.5, hjust = 0.5) +
  
  annotate("rect", xmin = 2.3, xmax = 3.7, ymin = 4.5, ymax = 5.5,
           fill = "#ffcccc", color = "#cc0000", size = 1.5) +
  annotate("text", x = 3, y = 5, label = "2. Spreads ↑\nStress", 
           fontface = "bold", size = 3.5, hjust = 0.5) +
  
  annotate("rect", xmin = 4.3, xmax = 5.7, ymin = 4.5, ymax = 5.5,
           fill = "#ffcccc", color = "#cc0000", size = 1.5) +
  annotate("text", x = 5, y = 5, label = "3. Credit\nSupply ↓", 
           fontface = "bold", size = 3.5, hjust = 0.5) +
  
  annotate("rect", xmin = 6.3, xmax = 7.7, ymin = 4.5, ymax = 5.5,
           fill = "#ffcccc", color = "#cc0000", size = 1.5) +
  annotate("text", x = 7, y = 5, label = "4. Demand ↓\nDisinflation", 
           fontface = "bold", size = 3.5, hjust = 0.5) +
  
  # Arrows connecting problem chain
  geom_segment(aes(x = 1.7, y = 5, xend = 2.3, yend = 5),
               arrow = arrow(length = unit(0.3, "cm")), color = "#cc0000", size = 1) +
  geom_segment(aes(x = 3.7, y = 5, xend = 4.3, yend = 5),
               arrow = arrow(length = unit(0.3, "cm")), color = "#cc0000", size = 1) +
  geom_segment(aes(x = 5.7, y = 5, xend = 6.3, yend = 5),
               arrow = arrow(length = unit(0.3, "cm")), color = "#cc0000", size = 1) +
  
  # Policy responses (blue boxes)
  annotate("rect", xmin = 0.3, xmax = 1.7, ymin = 2.5, ymax = 3.5,
           fill = "#ccccff", color = "#0000cc", size = 1.5) +
  annotate("text", x = 1, y = 3, label = "Policy:\nLiquidity", 
           fontface = "bold", size = 3, hjust = 0.5, color = "#0000cc") +
  
  annotate("rect", xmin = 2.3, xmax = 3.7, ymin = 2.5, ymax = 3.5,
           fill = "#ccccff", color = "#0000cc", size = 1.5) +
  annotate("text", x = 3, y = 3, label = "Policy:\nSwap Lines", 
           fontface = "bold", size = 3, hjust = 0.5, color = "#0000cc") +
  
  annotate("rect", xmin = 4.3, xmax = 5.7, ymin = 2.5, ymax = 3.5,
           fill = "#ccccff", color = "#0000cc", size = 1.5) +
  annotate("text", x = 5, y = 3, label = "Policy:\nCredit Ease", 
           fontface = "bold", size = 3, hjust = 0.5, color = "#0000cc") +
  
  annotate("rect", xmin = 6.3, xmax = 7.7, ymin = 2.5, ymax = 3.5,
           fill = "#ccccff", color = "#0000cc", size = 1.5) +
  annotate("text", x = 7, y = 3, label = "Policy:\nQE/LSAP", 
           fontface = "bold", size = 3, hjust = 0.5, color = "#0000cc") +
  
  # Arrows from problems to policies
  geom_segment(aes(x = 1, y = 4.5, xend = 1, yend = 3.5),
               arrow = arrow(length = unit(0.2, "cm")), color = "#2ca02c", 
               linetype = "dashed", size = 0.8, alpha = 0.6) +
  geom_segment(aes(x = 3, y = 4.5, xend = 3, yend = 3.5),
               arrow = arrow(length = unit(0.2, "cm")), color = "#2ca02c", 
               linetype = "dashed", size = 0.8, alpha = 0.6) +
  geom_segment(aes(x = 5, y = 4.5, xend = 5, yend = 3.5),
               arrow = arrow(length = unit(0.2, "cm")), color = "#2ca02c", 
               linetype = "dashed", size = 0.8, alpha = 0.6) +
  geom_segment(aes(x = 7, y = 4.5, xend = 7, yend = 3.5),
               arrow = arrow(length = unit(0.2, "cm")), color = "#2ca02c", 
               linetype = "dashed", size = 0.8, alpha = 0.6) +
  
  # Legend
  annotate("text", x = 0.5, y = 1.5, 
           label = "Red: Problem propagation chain\nBlue: Policy countermeasures (dashed arrows)",
           size = 3, hjust = 0, style = "italic") +
  
  theme_void() +
  xlim(0, 8) +
  ylim(1, 6) +
  coord_fixed(ratio = 1)
```

**[FIGURE 3 HERE] "Plumbing" diagram of crisis transmission breakdown and policy tools**

:::{.callout-note}
**Figure 3 Caption:** Flowchart showing: funding liquidity shock → interbank spreads/dealer stress → credit supply contraction → demand contraction → disinflation risk. Overlay policy tools: liquidity facilities, swap lines, QE/LSAP, forward guidance.
:::

## Balance-Sheet Policies and Unconventional Monetary Policy

Post-2008, central banks used their balance sheets as policy tools. A key analytical improvement is to distinguish:
1. **Liquidity provision** (steering money markets, bank funding)
2. **Credit easing/market functioning operations** (targeted backstops for specific dysfunctional markets)
3. **Large-scale asset purchases (QE/LSAPs)** (buying duration risk to compress longer-term yields)

### Bank of England: QE via Gilt Purchases Financed by Reserves

The UK's large-scale purchases began in March 2009 and were expanded repeatedly:
- **May 2009:** increased to **£125 billion**
- **August 2009:** increased to **£175 billion**
- **November 2009:** increased to **£200 billion**

Each expansion was financed by central bank reserves (not by borrowing or open-market operations).

The canonical BoE research estimate from this "first wave" is Joyce et al. (2010), which explicitly states that by February 2010 the MPC had made £200 billion of purchases (mostly gilts) and estimates QE may have depressed gilt yields by about **100 basis points**. The largest impact occurred via a **portfolio rebalancing channel**, though the authors acknowledge uncertainty about wider asset-price effects.

### Federal Reserve: LSAP1 and the Combination of Purchases and Credit Policies

The Fed's balance-sheet timeline records the core LSAP1 announcements:
- **November 25, 2008:** intention to purchase up to $100 billion agency debt and up to $500 billion agency MBS
- **March 18, 2009:** expansion to $1.25 trillion MBS, $200 billion agency debt, and $300 billion longer-term Treasuries

Research from the NY Federal Reserve finds LSAPs produced economically meaningful reductions in longer-term interest rates primarily through lower risk/term premia rather than only through expectations of future short rates.

### ECB: Bank-Based Nonstandard Measures, CBPP, and (by 2010) SMP

The ECB's 2009 nonstandard policy relied heavily on full-allotment liquidity and LTROs. It also introduced the **Covered Bond Purchase Programme (CBPP)** in May 2009, with purchases starting in July 2009 aimed at supporting a key bank funding segment. The program reached its intended nominal amount of **€60 billion** by June 30, 2010.

By May 2010, sovereign bond market dysfunction led the ECB to launch the SMP to support market depth and restore transmission while stating it would not alter the monetary-policy stance.

### A Compact "Channels" Summary for Undergraduates

A useful bridge between theory and practice is to show that unconventional policies work through recognizable transmission channels:

- **Portfolio balance:** central bank purchases reduce net supply of duration/risk to private portfolios → lower term premia → lower longer-term yields
- **Signaling/expectations management:** communication that rates will stay low "for an extended period" affects expected future short rates and risk premia
- **Market functioning:** targeted backstops reduce fire-sale spirals, improve liquidity, and reopen credit markets
- **International liquidity:** swap lines relieve currency-specific shortages and stabilize cross-border funding

```{r}
#| label: fig-balance-sheet
#| fig-cap: "Central Bank Balance Sheet Mechanics: Asset Purchase Financed by Reserves"
#| fig-height: 6
#| fig-width: 10

# Create stylized T-account representation
df_cb <- tribble(
  ~Type, ~Item, ~Before, ~After, ~x_pos, ~y_pos,
  "Assets", "Treasury/Gilts", "$100bn", "$300bn", 1, 3.5,
  "Assets", "MBS", "$50bn", "$250bn", 1, 2.5,
  "Assets", "Other", "$100bn", "$100bn", 1, 1.5,
  "Liabilities", "Reserves (banks)", "$400bn", "$600bn", 4, 3.5,
  "Liabilities", "Currency", "$100bn", "$100bn", 4, 2.5,
  "Liabilities", "Other", "$0bn", "$0bn", 4, 1.5
)

# Create background boxes for assets/liabilities
ggplot() +
  # Asset side
  annotate("rect", xmin = 0.3, xmax = 1.7, ymin = 0.5, ymax = 4.5,
           fill = "#ffcccc", color = "#cc0000", size = 1) +
  annotate("text", x = 1, y = 4.8, label = "CENTRAL BANK ASSETS",
           fontface = "bold", size = 4, hjust = 0.5) +
  # Liability side
  annotate("rect", xmin = 3.3, xmax = 4.7, ymin = 0.5, ymax = 4.5,
           fill = "#ccccff", color = "#0000cc", size = 1) +
  annotate("text", x = 4, y = 4.8, label = "CENTRAL BANK LIABILITIES",
           fontface = "bold", size = 4, hjust = 0.5) +
  # Asset items
  annotate("text", x = 0.5, y = 3.5, label = "Treasury/Gilts:",
           hjust = 0, fontface = "bold", size = 3) +
  annotate("text", x = 1.4, y = 3.5, label = "$100bn → $300bn",
           hjust = 1, size = 3, color = "#1f77b4") +
  annotate("text", x = 0.5, y = 2.5, label = "MBS:",
           hjust = 0, fontface = "bold", size = 3) +
  annotate("text", x = 1.4, y = 2.5, label = "$50bn → $250bn",
           hjust = 1, size = 3, color = "#1f77b4") +
  annotate("text", x = 0.5, y = 1.5, label = "Other:",
           hjust = 0, fontface = "bold", size = 3) +
  annotate("text", x = 1.4, y = 1.5, label = "$100bn",
           hjust = 1, size = 3) +
  # Liability items
  annotate("text", x = 3.5, y = 3.5, label = "Reserves (deposits):",
           hjust = 0, fontface = "bold", size = 3) +
  annotate("text", x = 4.4, y = 3.5, label = "$400bn → $600bn",
           hjust = 1, size = 3, color = "#ff7f0e") +
  annotate("text", x = 3.5, y = 2.5, label = "Currency:",
           hjust = 0, fontface = "bold", size = 3) +
  annotate("text", x = 4.4, y = 2.5, label = "$100bn",
           hjust = 1, size = 3) +
  # Annotations
  annotate("text", x = 2.5, y = 3.5, label = "↔", 
           size = 6, fontface = "bold", color = "#2ca02c") +
  annotate("text", x = 2.5, y = 0.2, 
           label = "Asset purchase ↑ → Reserves created ↑\nComposition: longer-duration/riskier assets ↑",
           size = 3, hjust = 0.5, style = "italic") +
  theme_void() +
  xlim(0, 5) +
  ylim(-0.2, 5)
```

**[FIGURE 4 HERE] Central bank balance sheet mechanics (stylized T-accounts)**

:::{.callout-note}
**Figure 4 Caption:** Asset purchase financed by reserves: central bank assets ↑ (gilts/MBS), liabilities ↑ (reserves). Emphasize that the purchase changes the composition of assets held by the private sector, and that reserves are liabilities of the central bank.
:::

```{r}
#| label: tbl-unconventional
#| echo: false

unconventional_matrix <- tribble(
  ~`Policy Type`, ~`Instrument Mechanics`, ~`Targeted Market Failure`, ~`Transmission Channels`,
  "Liquidity Provision", "Lend reserves at policy rate (unlimited supply)", "Interbank funding freeze, hoarding", "Lower interbank spreads, restore money market function",
  "Credit Easing/Facilities", "Backstop specific markets (CP, ABS, dealer repo)", "Fire sales in specific segments", "Remove dysfunction, restore pricing, reduce credit rationing",
  "QE/LSAP", "Purchase long-duration/risky assets with reserves", "Term premium elevation, portfolio imbalance", "Lower longer rates via portfolio rebalancing, expectations effects",
  "Forward Guidance", "Communicate future policy path", "Uncertainty about near-term/far-term rates", "Anchor expectations, shape term structure, signal commitment"
)

knitr::kable(unconventional_matrix, format = "html",
             caption = "Unconventional-Policy Classification: Mechanisms, Targets, Channels",
             escape = FALSE) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = TRUE) %>%
  kableExtra::column_spec(1, width = "12%") %>%
  kableExtra::column_spec(2:4, width = "29.3%")
```

**[TABLE 2 HERE] Unconventional-policy classification matrix**

:::{.callout-note}
**Table 2 Caption:** 
Rows: liquidity provision, credit easing, QE/LSAP, forward guidance
Columns: instrument mechanics, targeted market failure, main transmission channels, risks (moral hazard, fiscal exposure, exit complexity)
:::

## International Spillovers and Central Bank Coordination

The GFC highlighted international funding spillovers and the limits of purely domestic policy tools. The crisis blurred the distinction between monetary and financial stability and revived practical central bank cooperation. The Fed's documentation notes that the FOMC authorized swap lines with multiple foreign central banks over time and that these temporary arrangements expired on February 1, 2010.

From the ECB side, October 2008 press releases describe expanding collateral and enhancing liquidity provision, including the offer of US dollar liquidity through FX swaps—demonstrating that euro-area monetary operations became explicitly multi-currency during crisis management.

A useful empirical bridge for students is that BIS reporting at the time described dollar funding shortages in global banking and highlighted the interplay between central bank measures and interbank spread declines.

```{r}
#| label: fig-swaplines
#| fig-cap: "Central Bank Swap Line Network (2007–2010): International Dollar Liquidity Provision"
#| fig-height: 6
#| fig-width: 10

# Create swap line network
swap_lines_edges <- tribble(
  ~from, ~to, ~date_start, ~date_end, ~type,
  "Fed", "ECB", "2007-12", "2010-02", "Dec 2007",
  "Fed", "SNB", "2007-12", "2010-02", "Dec 2007",
  "Fed", "BoE", "2008-09", "2010-02", "Sep 2008",
  "Fed", "BoJ", "2008-12", "2010-02", "Dec 2008",
  "Fed", "Other CBs", "2008-12", "2010-02", "Expanded"
)

# Create nodes
swap_cbcenters <- tribble(
  ~bank, ~x, ~y, ~size,
  "Fed", 2.5, 2.5, 8,
  "ECB", 0.5, 4, 6,
  "BoE", 1.5, 0.5, 5,
  "SNB", 4.5, 4, 5,
  "BoJ", 4.5, 1, 5,
  "Other CBs", 1.5, 3.5, 4
)

# Plot
ggplot() +
  # Draw edges (swap lines)
  geom_segment(aes(x = 2.5, y = 2.5, xend = 0.5, yend = 4),
               arrow = arrow(length = unit(0.2, "cm"), type = "open"),
               color = "#1f77b4", size = 1, alpha = 0.7) +
  geom_segment(aes(x = 2.5, y = 2.5, xend = 1.5, yend = 0.5),
               arrow = arrow(length = unit(0.2, "cm"), type = "open"),
               color = "#1f77b4", size = 1, alpha = 0.7) +
  geom_segment(aes(x = 2.5, y = 2.5, xend = 4.5, yend = 4),
               arrow = arrow(length = unit(0.2, "cm"), type = "open"),
               color = "#1f77b4", size = 1, alpha = 0.7) +
  geom_segment(aes(x = 2.5, y = 2.5, xend = 4.5, yend = 1),
               arrow = arrow(length = unit(0.2, "cm"), type = "open"),
               color = "#1f77b4", size = 1, alpha = 0.7) +
  geom_segment(aes(x = 2.5, y = 2.5, xend = 1.5, yend = 3.5),
               arrow = arrow(length = unit(0.2, "cm"), type = "open"),
               color = "#ff7f0e", size = 1, alpha = 0.7) +
  # Draw nodes
  geom_point(data = swap_cbcenters, aes(x = x, y = y, size = size),
             color = "#d62728", alpha = 0.8) +
  # Labels
  geom_text(data = swap_cbcenters, aes(x = x, y = y, label = bank),
            fontface = "bold", size = 3.5, hjust = 0.5, vjust = 0.5) +
  # Date labels on connections
  annotate("text", x = 1.3, y = 3.3, label = "Dec 2007", size = 2.5, color = "#1f77b4") +
  annotate("text", x = 1.8, y = 2.0, label = "Sep 2008", size = 2.5, color = "#1f77b4") +
  annotate("text", x = 3.5, y = 3.3, label = "Dec 2007", size = 2.5, color = "#1f77b4") +
  annotate("text", x = 3.5, y = 1.8, label = "Dec 2008", size = 2.5, color = "#1f77b4") +
  annotate("text", x = 2.0, y = 3.0, label = "2008-09+", size = 2.5, color = "#ff7f0e") +
  # Legend
  annotate("text", x = 0.2, y = 0.2, 
           label = "All swap lines: Temporary, expired Feb 1, 2010\nPurpose: Provide USD liquidity during dollar shortage",
           size = 2.8, hjust = 0, vjust = 0, style = "italic") +
  theme_void() +
  xlim(-0.3, 5.2) +
  ylim(-0.2, 4.5) +
  coord_fixed() +
  labs(title = "Federal Reserve Dollar Swap Line Network (2007–2010)")
```

**[FIGURE 5 HERE] Swap-line network diagram (2007–2010)**

:::{.callout-note}
**Figure 5 Caption:** Simple network diagram showing Fed dollar swap lines with ECB/SNB (Dec 2007) and broader network as crisis intensified; annotate expiry Feb 2010.
:::

## Evaluation, Controversies, and the Post-2010 Legacy

A post-GFC chapter must teach not only what central banks did, but how economists and institutions evaluated these actions.

### Was Monetary Policy Effective During the Crisis?

Frederic Mishkin (2009) argues the view that monetary policy is ineffective during crises is wrong and risks policy inaction. He supports a "risk-management approach" where aggressive easing can reduce adverse feedback loops between financial distress and the macroeconomy. Mishkin (2011) frames the crisis as changing the perceived boundaries of monetary policy strategy, including the role of financial stability.

### How Large Were LSAP/QE Effects on Yields?

Institutional research is careful and typically focuses on financial variables (yields, spreads) more than precise GDP/inflation effects:

- **BoE estimates:** QE reduced gilt yields by about 100 basis points, with the portfolio balance channel important. Broader effects on output and inflation remain uncertain.
- **NY Fed evidence:** LSAPs reduced longer-term rates across many securities mainly via term/risk premia reductions.

A useful approach for 3rd-year students is to treat these as **partial-equilibrium financial effects** that plausibly feed into demand, rather than to claim a single definitive macro multiplier.

### Rules, Discretion, Moral Hazard, and Legitimacy

The GFC revived classical lender-of-last-resort concerns about moral hazard and the "rules vs discretion" problem, now under modern conditions where (i) solvency is hard to assess in real time and (ii) the shadow banking system is large. The debate echoes time-consistency and credibility themes: crisis interventions can be welfare-improving ex post but create incentives for risk-taking ex ante. Primary sources emphasize the need to design facilities with collateral and pricing rules to mitigate moral hazard.

### From 2010 Onward: How the Crisis "Rewrote Chapter 8"

Pre-crisis treatments of inflation control focused on policy rates, reserve supply, and (depending on the regime) money or exchange-rate targets. Post-2007, the toolkit expanded to include balance-sheet policies, market backstops, and international liquidity provision. The key legacy is that "inflation control" is now inseparable from the resilience of financial plumbing and the credibility of central bank frameworks under extreme stress.

```{r}
#| label: fig-toolkit
#| fig-cap: "Policy Toolkit Expansion: Central Banks Adding Instruments as Constraints Tightened (2007–2010)"
#| fig-height: 6
#| fig-width: 10

# Toolkit expansion stages
toolkit_stages <- tribble(
  ~Stage, ~`Policy Tool`, ~`Constraint Tightened`, ~`When Deployed`, ~x_pos, ~y_pos,
  1, "Conventional\nRate Cuts", "Rates ↓ as outlook weakens", "2007–2008", 1, 1,
  2, "Lender of\nLast Resort", "Policy rate → ZLB", "2007–2008", 2.5, 1,
  3, "Balance-Sheet\nPolicy (QE/LSAP)", "Rates at ZLB, demand ↓", "2009–2010", 4, 1,
  4, "International\nCoordination", "Dollar shortage, spillovers", "2007–2010", 5.5, 1,
  5, "Forward\nGuidance", "Managing expectations at ZLB", "2009–2010", 7, 1
)

# Create step chart
ggplot(toolkit_stages, aes(x = x_pos, y = 1)) +
  # Draw boxes
  geom_rect(aes(xmin = x_pos - 0.35, xmax = x_pos + 0.35, ymin = 0.7, ymax = 1.3),
            fill = "#1f77b4", alpha = 0.7, color = "#0000cc", size = 1) +
  # Stage numbers
  geom_text(aes(label = Stage), x = 0.4, y = c(1.1, 1.1, 1.1, 1.1, 1.1),
            fontface = "bold", size = 4, color = "white") +
  # Tool labels
  geom_text(aes(label = `Policy Tool`, y = 1), 
            fontface = "bold", size = 3, hjust = 0.5, vjust = 0.5) +
  # Connecting arrows
  geom_segment(aes(x = 1.35, y = 1, xend = 2.15, yend = 1),
               arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
               color = "#ff7f0e", size = 1.5) +
  geom_segment(aes(x = 2.85, y = 1, xend = 3.65, yend = 1),
               arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
               color = "#ff7f0e", size = 1.5) +
  geom_segment(aes(x = 4.35, y = 1, xend = 5.15, yend = 1),
               arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
               color = "#ff7f0e", size = 1.5) +
  geom_segment(aes(x = 5.85, y = 1, xend = 6.65, yend = 1),
               arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
               color = "#ff7f0e", size = 1.5) +
  # Constraint labels below
  annotate("text", x = 1, y = 0.45, label = "Outlook\nWorsens", size = 2.5, hjust = 0.5) +
  annotate("text", x = 2.5, y = 0.45, label = "Rates →\nZLB", size = 2.5, hjust = 0.5) +
  annotate("text", x = 4, y = 0.45, label = "Demand ↓\nZLB Binding", size = 2.5, hjust = 0.5) +
  annotate("text", x = 5.5, y = 0.45, label = "Global\nShortages", size = 2.5, hjust = 0.5) +
  annotate("text", x = 7, y = 0.45, label = "Inflation\nHoping", size = 2.5, hjust = 0.5) +
  # Title callout
  annotate("rect", xmin = 0.2, xmax = 7.8, ymin = 1.6, ymax = 1.85,
           fill = "#ffffcc", color = "#ffaa00", size = 1, alpha = 0.9) +
  annotate("text", x = 4, y = 1.725, 
           label = "Progressive Crisis Response: Each tool deployed as prior constraints tightened",
           size = 3.5, fontface = "bold", hjust = 0.5) +
  theme_void() +
  coord_fixed(xlim = c(0, 8), ylim = c(0.2, 2))
```

**[FIGURE 6 HERE] "Policy toolkit expansion" schematic**

:::{.callout-note}
**Figure 6 Caption:** Schematic ladder showing how central banks added instruments as constraints tightened: conventional rate policy → lender-of-last-resort → balance-sheet policy → coordination.
:::

## Summary and Key Terms

### Summary

Between 2007 and 2010, central banks responded to a breakdown in monetary transmission by combining:

- **Rapid conventional rate cuts** to (or near) the effective lower bound
- **Large-scale liquidity provision and emergency facilities** to keep key markets functioning (interbank, repo/dealer funding, commercial paper, ABS)
- **International swap-line networks** to relieve global dollar shortages
- **Asset purchases/QE/LSAPs** to compress longer-term yields and support demand when rates were constrained

The episode permanently expanded what "monetary policy" means in practice: operational capacity, balance-sheet tools, and coordination are now part of the standard toolkit discussion, even if not used in normal times.

### Key Terms

- **Effective Lower Bound (ELB):** practical lower bound on policy rates given institutional constraints
- **Lender of Last Resort (LOLR):** central bank liquidity support to solvent but illiquid institutions/markets, designed to prevent runs and fire sales
- **Fixed Rate Full Allotment:** central bank supplies demanded reserves/liquidity at a fixed policy rate (ECB, Oct 2008)
- **Quantitative Easing (QE)/LSAP:** large-scale asset purchase programs aimed at lowering longer-term yields and supporting spending
- **Swap Lines:** reciprocal currency arrangements to provide foreign-currency liquidity (notably dollars) to overseas markets

### Review Questions

1. Explain why the policy rate alone could not restore normal transmission in late 2008. Use at least two examples of market dysfunction (e.g., commercial paper, dealer funding, interbank markets).

2. Compare the BoE's SLS and QE programs: which balance-sheet problems did each target, and how did their mechanics differ?

3. Define "fixed rate full allotment" and explain why the ECB adopted it in October 2008.

4. Using BoE and NY Fed evidence, explain the portfolio balance channel of QE/LSAPs and how it differs from pure "signaling."

5. What was the stated intended purpose of the ECB SMP in May 2010? Why did the ECB emphasize it would not change the stance of policy?

6. Discuss the trade-off between crisis intervention and moral hazard using one facility (PDCF, CPFF, or SLS) as your example.

## References
Primary references for the GFC-period policy chronology and crisis tools in this chapter include BIS working papers, central bank press releases/minutes, and summary research studies [@allenMoessner2010BIS310; @borioDisyatat2009UMPAppraisal; @boe2007NorthernRockLiquiditySupport; @boe2008SpecialLiquidityScheme; @boe2009AssetPurchaseFacilityJan; @boe2009MPCMinutesMar; @boe2009MPCAnnouncement; @boe2009APFMayIncrease; @boe2009APFAugIncrease; @boe2009APFNovIncrease; @joyceEtAl2010QEWP393; @ecb2008TenderProcedureCorridor; @ecb2008CollateralFrameworkLiquidity; @ecb2009MonetaryPolicyDecisionsMay; @ecb2010CBPPCompleted; @ecb2010SMPMeasures; @frb2007CoordinatedLiquidityMeasures; @frb2008CPFFCreation; @fomc2008MinutesDec1516; @fomc2009MinutesMar1718; @mishkin2009NBER14678; @mishkin2011NBER16755].

::: {#refs}
:::
